---
title: "Scotland segmented over 75"
output:
  word_document: default
  html_notebook: default
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Introduction

To undertake segmented regression (one and two break) for Scotland age group over 75 years. Employs full data range available (Q1 1990-Q2 2018)

* Load packages
* Load data
* Ensure correctly labelled
* Convert dates to identifiable quarters
* Limit over 75
* Generate linear model
* Generate segmented 
* Run Davies test

## Load packages

```{r}
#install.packages("pacman")
pacman::p_load(
  tidyverse,
  segmented,
  plotly,
  dplyr,
  readr,
  ggplot2,
  broom,
  tidyr
               )
```

## Load data and assign to object "data"

```{r}
data <- read.csv("Data/QuarterlyASMR1990_2018_OVER75.csv")
data
```

#Tidy periods into quarters, select all age variables, filter to M&F only
```{r}

tidied_OVER75rates <-
  data %>%
  separate(Quarter, into = c("year","quarter"), sep = 4) %>% 
  mutate(
   year = as.double(year),
   quarter = as.double(quarter)
  ) %>% 
  mutate(
    time = year + (quarter / 4) - 0.125
  ) %>% 
  select(time, Sex, Rate)%>%
  rename(sex=Sex)


tidied_OVER75rates

```

# plot mortality rates by time
```{r}
plot<-ggplot(tidied_OVER75rates, aes(x=time,y=Rate, color=sex))+geom_line(size=1.0001, show.legend = TRUE)+theme(text=element_text(size=14))
plot
```

# Segmented regression 
##Generate linear model for both sexes

```{r}
tidied_OVER75rates %>% 
  filter(sex == "M") -> O75male
  lm(Rate ~ time, data = O75male) -> lmodO75_male

tidied_OVER75rates %>%
  filter(sex == "F") -> O75female
  lm(Rate ~ time, data =  O75female) -> lmodO75_female

```

##View linear model 

```{r}
summary (lmodO75_male)
summary (lmodO75_female)

```


## Davies test 
```{r}
davies_O75male = davies.test(lmodO75_male, ~time, k=111)
davies_O75female = davies.test(lmodO75_female, ~time, k=111)
davies_O75male
davies_O75female
```
  
## Get trend line of linear model fit to whole timeseries 
  

```{r}
my.coeftrendlineO75male <- coef(lmodO75_male)
my.coeftrendlineO75female <- coef(lmodO75_female)
addtrendline <- plot + geom_abline(intercept = my.coeftrendlineO75male[1], col="Blue", size=1.0001,
                                       slope = my.coeftrendlineO75male[2], 
                                       aes(colour = "Overall"))+
  geom_abline(intercept = my.coeftrendlineO75female[1], col="Blue", size=1.0001,
                                       slope = my.coeftrendlineO75female[2], 
                                       aes(colour = "Overall"))
addtrendline
```
## Segmented regression function - one break point

```{r}
 my.seg_male_O75<-segmented(  lmodO75_male, seg_rate.Z = ~time, psi = NA, control = seg.control(K=1))
 my.seg_female_O75<-segmented(  lmodO75_female, seg_rate.Z = ~time, psi = NA, control = seg.control(K=1))

```
  
## outputs from segmented function

```{r}
summary(my.seg_male_O75)
summary(my.seg_female_O75)
my.fitted_male_O75 <- fitted(my.seg_male_O75)
my.fitted_female_O75 <- fitted(my.seg_female_O75)

```
#Extract breakpoint and CI

```{r}
breakpointmale_O75 <- my.seg_male_O75$psi[, 2]
breakpointfemale_O75 <- my.seg_female_O75$psi[, 2]
breakpoint_SEmale_O75 <- my.seg_male_O75$psi[, 3]
breakpoint_SEfemale_O75 <- my.seg_female_O75$psi[, 3]
breakpoint_low_CImale_O75 <- breakpointmale_O75 - (breakpoint_SEmale_O75*1.96)
breakpoint_up_CImale_O75 <- breakpointmale_O75 + (breakpoint_SEmale_O75*1.96)
breakpoint_low_CIfemale_O75 <- breakpointfemale_O75 - (breakpoint_SEfemale_O75*1.96)
breakpoint_up_CIfemale_O75 <- breakpointfemale_O75 + (breakpoint_SEfemale_O75*1.96)

breakpointmale_O75
breakpoint_low_CImale_O75
breakpoint_up_CImale_O75
breakpointfemale_O75
breakpoint_low_CIfemale_O75
breakpoint_up_CIfemale_O75
```  
#Plot model
```{r}
tidied_OVER75rates %>% 
  filter(sex =="M") %>% 
  ggplot(
    aes(x = time)
  ) + 
  geom_point(
    aes(y = Rate)
  ) + 
  geom_line(
    aes(y = my.fitted_male_O75)
  )
tidied_OVER75rates %>% 
  filter(sex =="F") %>% 
  ggplot(
    aes(x = time)
  ) + 
  geom_point(
    aes(y = Rate)
  ) + 
  geom_line(
    aes(y = my.fitted_female_O75) 
  )
```
#Model predicted values to data frame tidied_allagerates
```{r}
tidied_OVER75rates %>%
  filter(sex == "M") %>% 
  mutate(
    onebreak_fit_O75 = fitted(
      my.seg_male_O75 
     
      )
    ) -> malefitted_O75

tidied_OVER75rates %>%
  filter(sex == "F") %>% 
  mutate(
    onebreak_fit_O75 = fitted(
      my.seg_female_O75 

      )
    ) -> femalefitted_O75
plot(malefitted_O75$onebreak_fit_O75)
plot(femalefitted_O75$onebreak_fit_O75)
```

##Combine male and female observed and predicted

```{r}
femalefitted_O75 %>% 
  select(
    time, sex, observed = Rate,
    predicted = onebreak_fit_O75
  ) %>% 
  bind_rows(
    malefitted_O75 %>% 
      select(
        time, sex, observed = Rate,
        predicted = onebreak_fit_O75
      ) 
  ) ->obs_fit_both_O75

age <- "over75"
obs_fit_both_O75$agegroup = age

obs_fit_both_O75
```
#Plot obs, predicted, and lm
```{r}
obs_fit_both_O75 %>% 
  ggplot(
    aes (x=time, color=sex)
  ) +
  geom_point(
    aes (y=observed)) +
  geom_line ( 
    aes (y=predicted)) +
  stat_smooth(method="lm",se=F,linetype="dashed",
    aes (y=observed)
  )
```

##Getsegmented regression coefficients
```{r}
coef(my.seg_male_O75)
coef(my.seg_female_O75)
```
##Plot using coefficients and geom_abline
```{r}
 # Get coeffiencents male
  b0_aspec_m_O75 <- coef(my.seg_male_O75)[[1]]
  b1_aspec_m_O75 <- coef(my.seg_male_O75)[[2]]
  c1_aspec_m_O75 <- coef(my.seg_male_O75)[[2]] + coef(my.seg_male_O75)[[3]]
  
  break1_aspec_m_O75 <- my.seg_male_O75$psi[[2]]
  
  # Get coefficients female
    b0_aspec_f_O75 <- coef(my.seg_female_O75)[[1]]
  b1_aspec_f_O75 <- coef(my.seg_female_O75)[[2]]
  c1_aspec_f_O75 <- coef(my.seg_female_O75)[[2]] + coef(my.seg_female_O75)[[3]]
  
  break1_aspec_f_O75 <- my.seg_female_O75$psi[[2]]
  
# plot
 p <- obs_fit_both_O75 %>% 
  ggplot(
    aes (x=time, color=sex)
  ) +
  geom_point(
    aes (y=observed), alpha=0.4) +
  geom_line (
    aes (y=predicted), size=1) +
  geom_abline(intercept = b0_aspec_m_O75, slope = b1_aspec_m_O75, color="#00BFC4", linetype="dashed") +
    geom_abline(intercept = b0_aspec_f_O75, slope = b1_aspec_f_O75, color="#F8766D", linetype= "dashed") +
    ylim(0,15000)+
    labs(x="Year", y="Age-standardised mortality rate", title = 
        "Age-standardised quarterly rolling annual mortality rates 
         and fitted segmented model, males and females, over 75 years, 
         Scotland, 1990 to 2018") +
    scale_x_continuous (breaks=c(1990, 1994, 1998, 2002, 2006, 2010, 2014, 2018))+
    geom_vline(xintercept = my.seg_male_O75[["psi"]][[2]], color = "#00BFC4", linetype="dotdash")+
    geom_vline(xintercept = my.seg_female_O75[["psi"]][[2]], color = "#F8766D", linetype="dotdash") +
   annotate(geom="text", x=2012, y=2500, label="Male 75+\nbreakpoint\nApr 2013-\n Mar 2014 ")+
   annotate(geom="text", x=2017, y=2500, label="Female 75+\nbreakpoint\nOct 2013-\nSep 2014")
   
 
 plot(p)

 
    ggsave (
    "Figures/Segmentedplot_Scotland_75+.png", height=15, width=20, units="cm")

```



##2 breakpoint segmented 
```{r}
my.seg_male2_O75 <-segmented( lmodO75_male, seg_rate.Z = ~timepoint, psi = NA, control = seg.control(K=2))
my.seg_female2_O75<- segmented(lmodO75_female, seg_rate.Z = ~timepoint, psi = NA, control = seg.control(K=2))

summary(my.seg_male2_O75)
summary(my.seg_female2_O75)
```



##AIC and BIC to compare one and two break models
```{r}
AIC(my.seg_male_O75, my.seg_male2_O75)
AIC(my.seg_female_O75, my.seg_female2_O75)
BIC(my.seg_male_O75, my.seg_male2_O75)
BIC(my.seg_female_O75, my.seg_female2_O75)
```

#Extract breakpoint and CI for 2 break model

```{r}



breakpointmale_O75_2break <- my.seg_male2_O75$psi[, 2]
breakpointfemale_O75_2break <- my.seg_female2_O75$psi[, 2]
breakpoint_SEmale_O75_2break <- my.seg_male2_O75$psi[, 3]
breakpoint_SEfemale_O75_2break <- my.seg_female2_O75$psi[, 3]
breakpoint_low_CImale_O75_2break <- breakpointmale_O75_2break - (breakpoint_SEmale_O75_2break*1.96)
breakpoint_up_CImale_O75_2break <- breakpointmale_O75_2break + (breakpoint_SEmale_O75_2break*1.96)
breakpoint_low_CIfemale_O75_2break <- breakpointfemale_O75_2break - (breakpoint_SEfemale_O75_2break*1.96)
breakpoint_up_CIfemale_O75_2break <- breakpointfemale_O75_2break + (breakpoint_SEfemale_O75_2break*1.96)

breakpointmale_O75_2break
breakpoint_low_CImale_O75_2break
breakpoint_up_CImale_O75_2break
breakpointfemale_O75_2break
breakpoint_low_CIfemale_O75_2break
breakpoint_up_CIfemale_O75_2break
``` 




And now to save these values 


```{r}
save(
  breakpointmale_O75_2break, 
  breakpointfemale_O75_2break,
  breakpoint_SEmale_O75_2break,
  breakpoint_SEfemale_O75_2break,
  breakpoint_low_CImale_O75_2break, 
  breakpoint_up_CImale_O75_2break,
  breakpoint_low_CIfemale_O75_2break,
  breakpoint_up_CIfemale_O75_2break,
  breakpointmale_O75,
  breakpoint_low_CImale_O75,
  breakpoint_up_CImale_O75,
  breakpointfemale_O75,
  breakpoint_low_CIfemale_O75,
  breakpoint_up_CIfemale_O75,
file = "Data/summary_stats_75andover.rdata"
)

```








